---
title: "Visualization in R"
output: html_document
date: "April 9, 2024"
---

**Install the `ggplot2` and `RColorBrewer` packages:**

```r
install.packages(c('ggplot2', 'RColorBrewer'))
```

```{r}
library(dplyr, warn.conflicts = F)
library(ggplot2)
library(RColorBrewer)

npp <- read.csv("http://files.ntsg.umt.edu/data/GIS_Programming/data/GPPDI_ClassA_NPP_162_R2.csv")
npp[npp$TEMP_ANN < -900,'TEMP_ANN'] = NA
npp[npp$PREC_ANN < -900,'PREC_ANN'] = NA
npp[npp$BNPP_C < -900,'BNPP_C'] = NA

species <- read.csv("http://files.ntsg.umt.edu/data/GIS_Programming/data/species.csv")
plots <- read.csv("http://files.ntsg.umt.edu/data/GIS_Programming/data/plots.csv")
surveys <- read.csv("http://files.ntsg.umt.edu/data/GIS_Programming/data/surveys.csv")

df.surveys <- surveys %>%
  left_join(species, by = 'species_id') %>%
  left_join(plots, by = 'plot_id')
```
## `ggplot2` and The Grammar of Graphics
What is ggplot2?

```{r}
df.surveys %>%
  head

summary(df.surveys)

ggplot(df.surveys, # This is the data you want to work with
       mapping = aes(x = weight, # This is how you specifiy the axes
                     y = hindfoot_length)) +
  geom_point() # This is how you define the type of data plotting (add layers)

```

The '+' is syntax for adding layers to a plot. We start with a definition of a plot, then use + to add layers to the layout

We can also create a template for a plot:

```{r}
base <- ggplot(df.surveys,
               mapping = aes(x = weight, y = hindfoot_length))

class(base)

# This is the same code as what we plotted above
base + geom_point()
```
We can control aspects of the styling per layer within the layer itself:
We can also save plots as images:

```{r}

base + geom_point(alpha = 0.1, color = 'orchid') # We can adjust the alpha to see areas with greater data density

# Mapping where colors are determined from another column (e.g. species)
base +
  geom_point(aes(color = species_id), alpha = 0.1)

# Saving plots
ggsave(width = 5, height = 4, dpi = 172,
       file = 'my_plot.png')

```

Challenge: create a plot where the points are colored by plot type, and set the "shape" aesthetic to the "sex" column

```{r}

base + geom_point(aes(color = plot_type, shape = sex), alpha = 0.1)

```
Plotting time series data:

We want to know how many rodents we captured each year.

```{r}

annual.counts <-          df.surveys %>%
                              group_by(year, genus) %>% # Grouping by multiple columns is going to show values for every genus for each year
                              filter(taxa == 'Rodent') %>%
                              tally() 

annual.counts <- df.surveys %>%
  filter(taxa == 'Rodent') %>%
  group_by(year, genus) %>%
  summarize(count = n())

annual.counts

```

Let's tell ggplot how to separate the data by genus

```{r}

ggplot(annual.counts, mapping = aes(x = year, y = count, color = genus)) +
  geom_line()

```

Changing geometries

```{r}

ggplot(annual.counts, mapping = aes(x = year, y = count, color = genus)) +
  geom_bar()

# We get an error! Why?
base1 <- ggplot(annual.counts, mapping = aes(x = year, y = count, fill = genus))

base1 + geom_bar(stat = 'identity')

# It stacks automatically. Here's what to do when we want to remove the stacking
base1 + geom_bar(stat = 'identity', position = 'dodge')

# Multiplotting / Faceting
base1 +
  geom_bar(stat = 'identity', position = 'dodge') +
  facet_wrap(~ genus) # This is how you segment subplots

# Facet Grid

```

Coordinate systems and multiple layers
Look at relationship between aboveground NPP and belowground NPP

```{r}

ggplot(npp, aes(x = ANPP_C, y = BNPP_C)) +
  geom_point()

ggplot(npp, aes(x = ANPP_C, y = BNPP_C)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = 'steelblue',
              linetype = 'dashed') +
  coord_equal(xlim = c(0,1000), ylim = c(0,1000)) # Specifies that we want to use an equal area coordinate system to force the plot to be square

# Below, the geom_smooth function is too smooth, so we want to give it some other options
ggplot(npp, aes(x = TEMP_ANN, y = TNPP_C)) +
  geom_point() +
  geom_smooth()

ggplot(npp, aes(x = TEMP_ANN, y = TNPP_C)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  geom_vline(xintercept = 0)

# Customizing plots
ggplot(npp, aes(x = TEMP_ANN, y = TNPP_C,
                color = BIOMENEW)) +
  geom_point() +
  geom_smooth(method = 'lm')

ggplot(npp, aes(x = TEMP_ANN, y = TNPP_C)) +
  geom_point(aes(color = BIOMENEW)) +
  geom_smooth(method = 'lm')



```